\\LARGE Decision-Optimal Probabilistic Forecast Reconciliation for Expected Shortfall in Hierarchical Time Series

### 2025-08-27 phd_application/phd_propossal.Rmd
update methodology sec

These tail risks are, however, not managed in isolation. Consider a hedge fund's investment portfolio returns, which compose of returns of different asset classes, and each asset class further contain returns of various individual securities. This is an example of a hierarchical structure, with multiple *aggregation levels*. Risk managers need to assess and forecast the returns or risk quantities at several levels, from the overall portfolio, down to individual securities. Ignoring this naturally-formed hierarchical structure can lead to misaligned forecasts, for example, the predicted returns of individual securities are low while the overall portfolio returns are high. This results in misallocation of capital and suboptimal risk management strategies.

@@

\color{red}
Importantly, ES behaves well under aggregation (satisfying subadditivity), making it a natural target when risks are rolled up through a hierarchy. This context motivates methods that deliver coherent, multi-level forecasts whose uncertainty aligns with the tail-risk metrics managers actually use.

(should i remove this?)
\color{black}

@@

Modern risk management and operations need more than a typical mean or median value--they need quantities from the full predictive distribution that capture uncertainty and tail behaviour, because many decisions are tail-sensitive (e.g., extreme demand in retail settings or energy generation shortfalls). While VaR is conceptually straightforward, it ignores the severity of losses beyond the VaR threshold and does not satisfy the subadditivity property\footnote{\textbf{Subadditivity.} A risk measure $\rho$ on loss variables is subadditive if $\rho(X+Y)\le \rho(X)+\rho(Y)$ for any losses $X,Y$. Expected Shortfall satisfies subadditivity for all $\alpha\in(0,1)$, whereas VaR can violate it. Thus, the portfolio ES cannot exceed the sums of individual security ESs.} [@Yamai2005-me]. Violating subadditivity is a critical problem, because it means that the risk measure of a combined portfolio can exceed the sum of stand-alone risks, which contradicts the principle of diversification. VaR therefore can mislead risk assessments across different levels of the hierarchy, and may underestimate the true risk exposure under market stress. ES, on the other hand, is a consistent metric that captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable measure for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii]. If ES is the metric that risk managers actually use, reconciliation methods should guarantee the coherence and optimality with respect to ES.

Most reconciliation methodology has been developed for point forecasts and are optimised for point scoring functions such as mean squared error (MSE). Generalising these methods to reconcile probabilistic forecasts does not guarantee optimality for tail-risk measures such as VaR or ES. Although @Panagiotelis2023-fs introduced score optimisation for reconciliation weights using proper scoring rules (energy score and variogram score) for full multivariate predictive distributions, they do not directly address the tail functionals. This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for tail-risk measures such as ES?"*

<!-- \color{red}
**AP: I have thought a bit about how we discuss the point v probabilistic issue. Ultimately, a VaR or ES forecast is a point forecast. However, a) any point forecast can be derived from a distributional forecast and b) scoring rules give us a way to obtain the optimal VaR or ES if we know the underlying probabilistic distribution. These two facts taken together mean we can naturally adapt the score optimisation framework of Panagiotelis et al (2023) to obtain optimal ES forecasts while also guaranteeing that these forecast come from an underlying probabilitistic forecast that is coherent.**
\color{black} -->

@@

As a result, the objective of my thesis is to bring these theoretical advances into the reconciliation framework and develop methods that produce coherent probabilistic forecasts optimised for ES. The goal is a reliable, consistent, and decision-aligned risk-forecasting framework that helps national and international organisations improve ES backtesting, stabilise capital, and allocate resources more effectively across hierarchical levels.

@@

@@

<!-- 
\tilde{\boldsymbol{y}}
\hat{\boldsymbol{y}}
-->

Base forecasts $\hat{\boldsymbol{y}}$, which we generate independently for each series, are typically *incoherent* (i.e. $\boldsymbol C \hat{\boldsymbol{y}} \neq 0$), while reconciled forecasts $\tilde{\boldsymbol{y}}$ are coherent by construction (i.e. $\boldsymbol C \hat{\boldsymbol{y}} = 0$). We obtain coherence via a linear mapping $\tilde{\boldsymbol{y}} = \boldsymbol P \hat{\boldsymbol{y}}$, where $\boldsymbol P$ maps the full *incoherent* space $\mathbb{R}^K$ to the linear *coherent* subspace $\mathbb{S} \subset \mathbb{R}^K$, where aggregation constraints hold.

<!-- 
This mapping $\boldsymbol P$ can be a projection operator $\boldsymbol S (\boldsymbol S' \boldsymbol S)^{-1} \boldsymbol S'$ [@Hyndman2011-jv], where $\boldsymbol S$ is a matrix designed such that $\boldsymbol S \boldsymbol C = 0$. The mapping $\boldsymbol P$ is not unique, and selecting an optimal $\boldsymbol P$ is the core design problem. In this project, we will focus solely on linear reconciliation.

@@

@@

For a random variable $Y$, the Expected Shortfall (ES) at level $\alpha \in (0, 1)$ is defined as $\mathrm{ES}_\alpha(Y) = E[Y \mid Y < \mathrm{VaR}_\alpha(Y)]$, where $\mathrm{VaR}_\alpha(Y)$ is the quantile $q$ at level $\alpha$. The ES alone is not elicitable, meaning there is no scoring rule or loss function that can uniquely identify the ES as the optimal optimiser. However, @Fissler2015-uu showed that the quantile function and ES are jointly elicitable, minimising the Fissler-Ziegel class of loss functions in expectation. One of which takes the form:

>>>>

These tail risks are, however, not managed in isolation. Consider a hedge fund's investment portfolio returns, which compose of returns of different asset classes, and each asset class further contain returns of various individual securities. This is an example of a hierarchical structure, with multiple *aggregation levels*. Risk managers need to assess and forecast the returns or risk quantities at several levels, from the overall portfolio, down to individual securities. Ignoring this naturally formed hierarchical structure can lead to misaligned forecasts, for example, the predicted returns of individual securities are low while the overall portfolio returns are high. This results in misallocation of capital and suboptimal risk management strategies.

@@

<!-- \color{red}
Importantly, ES behaves well under aggregation (satisfying subadditivity), making it a natural target when risks are rolled up through a hierarchy. This context motivates methods that deliver coherent, multi-level forecasts whose uncertainty aligns with the tail-risk metrics managers actually use.

(should i remove this?)
\color{black} -->

@@

Modern risk management and operations need more than a typical mean or median value--they need quantities from the full predictive distribution that capture uncertainty and tail behaviour, because many decisions are tail-sensitive (e.g., extreme demand in retail settings or energy generation shortfalls). While VaR is conceptually straightforward, it ignores the severity of losses beyond the VaR threshold and does not satisfy the subadditivity property\footnote{\textbf{Subadditivity.} A risk measure $\rho$ on loss variables is subadditive if $\rho(X+Y)\le \rho(X)+\rho(Y)$ for any losses $X,Y$. Expected Shortfall satisfies subadditivity for all $\alpha\in(0,1)$, whereas VaR can violate it. Thus, the portfolio ES cannot exceed the sums of individual security ESs.} [@Yamai2005-me]. Violating subadditivity is a critical problem, because it means that the risk measure of a combined portfolio can exceed the sum of stand-alone risks, which contradicts the principle of diversification. VaR therefore can mislead risk assessments across different levels of the hierarchy, and may underestimate the true risk exposure under market stress. ES, on the other hand, is a consistent metric that captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable measure for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii]. Although ES is the metric that risk managers actually use, there is a lack of research on forecasting ES within the reconciliation framework.

\color{red}
Most reconciliation methodology has been developed for point forecasts and are optimised for point scoring functions such as mean squared error (MSE). Generalising these methods to reconcile probabilistic forecasts does not guarantee optimality for tail-risk measures such as VaR or ES. Although @Panagiotelis2023-fs introduced score optimisation for reconciliation weights using proper scoring rules (energy score and variogram score) for full multivariate predictive distributions, they do not directly address the tail functionals. This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for tail-risk measures such as ES?"*
\color{black}

<!-- 
**AP: I have thought a bit about how we discuss the point v probabilistic issue. Ultimately, a VaR or ES forecast is a point forecast. However, a) any point forecast can be derived from a distributional forecast and b) scoring rules give us a way to obtain the optimal VaR or ES if we know the underlying probabilistic distribution. These two facts taken together mean we can naturally adapt the score optimisation framework of Panagiotelis et al (2023) to obtain optimal ES forecasts while also guaranteeing that these forecast come from an underlying probabilitistic forecast that is coherent.**
-->

@@

As a result, the objective of this thesis is to bring these theoretical advances into the reconciliation framework and develop methods that produce coherent probabilistic forecasts optimised for ES. The goal is a reliable, consistent, and decision-aligned risk-forecasting framework that helps national and international organisations improve ES backtesting, stabilise capital, and allocate resources more effectively across hierarchical levels.

@@

<!-- 
=========== OLD version

@@

\tilde{\boldsymbol{y}}
\hat{\boldsymbol{y}}

Base forecasts $\hat{\boldsymbol{y}}$, which we generate for each series, are typically *incoherent* (i.e. $\boldsymbol C \hat{\boldsymbol{y}} \neq 0$), while reconciled forecasts $\tilde{\boldsymbol{y}}$ are coherent by construction (i.e. $\boldsymbol C \hat{\boldsymbol{y}} = 0$). We obtain coherence via a linear mapping $\tilde{\boldsymbol{y}} = \boldsymbol P \hat{\boldsymbol{y}}$, where $\boldsymbol P$ maps the full *incoherent* space $\mathbb{R}^K$ to the linear *coherent* subspace $\mathbb{S} \subset \mathbb{R}^K$, where aggregation constraints hold.

This mapping $\boldsymbol P$ can be a projection operator $\boldsymbol S (\boldsymbol S' \boldsymbol S)^{-1} \boldsymbol S'$ [@Hyndman2011-jv], where $\boldsymbol S$ is a matrix designed such that $\boldsymbol S \boldsymbol C = 0$. The mapping $\boldsymbol P$ is not unique, and selecting an optimal $\boldsymbol P$ is the core design problem. In this project, we will focus solely on linear reconciliation.

@@

Let $\boldsymbol{y}_t \in \mathbb{R}^K$ denote a vector of $K$ time series variables from the hierarchy, at time $t$. For simplicity, we suppress the subscript $t$ and denote it as $\boldsymbol{y}$. These variables are subject to linear aggregation constraints. Base forecasts $\hat{\boldsymbol{y}}$, which we generate for each series, are typically *incoherent*, while reconciled forecasts $\tilde{\boldsymbol{y}}$ are coherent by construction. We obtain coherence via a linear mapping $\tilde{\boldsymbol{y}} = \boldsymbol P \hat{\boldsymbol{y}}$, where $\boldsymbol P$ maps the full *incoherent* space $\mathbb{R}^K$ to the linear *coherent* subspace $\mathbb{S} \subset \mathbb{R}^K$, where aggregation constraints hold. Ultimately, selecting an optimal $\boldsymbol P$ is the core design problem. In this project, we will focus solely on linear reconciliation.

For a random variable $Y$, the Expected Shortfall (ES) at level $\alpha \in (0, 1)$ is defined as $\mathrm{ES}_\alpha(Y) = E[Y \mid Y < \mathrm{VaR}_\alpha(Y)]$, where $\mathrm{VaR}_\alpha(Y)$ is the quantile $q$ at level $\alpha$. Thus, we can define $\widetilde{\mathrm{ES}}_k(\boldsymbol{P}) = \mathrm{ES}_\alpha(\tilde y_k(\boldsymbol{P}))$ as the ES of the $k$-th series after reconciliation, where $\tilde y_k(\boldsymbol{P})$ is the $k$-th element of the reconciled forecasts $\tilde{\boldsymbol{y}}(\boldsymbol{P}) = \boldsymbol P \hat{\boldsymbol{y}}$; and similarly define $\tilde q_k(\boldsymbol{P}) = \mathrm{VaR}_\alpha(\tilde y_k(\boldsymbol{P}))$ as the quantile of the $k$-th series after reconciliation.

@@

@Fissler2015-uu showed that the quantile function and ES are jointly elicitable, minimising the Fissler-Ziegel class of loss functions in expectation. One of which takes the form:
